Impact of Satellite Constellations on the Legacy Survey of Space and Time

### 2022-10-14 ms.tex
Updates from Overleaf

XXX--bottom line. Even in the case of multiple mega constellations in orbit, we expect Rubin would only lose $\sim$0.02\% of science pixels to streaks. Given reasonable satellite orbit predictions we find we can reduce pixel losses by a factor of 2 with minimal impact to the survey science goals. Even so, the tiny fraction of pixels lost makes mitigation strategies unnecessary. Twilight observations at high airmass with no scheduling mitigation could see pixel losses up to 0.06\%. 
\end{abstract}

@@

To accomplish this, there is a complex scheduler algorithm that takes into account various science priorities, including slew time, survey uniformity, and image depth \citep{naghib19}. The Rubin Scheduler algorithm creates weightings for various science priorities and uses the weighting to create a path for pointing observation.

However, one challenge for the LSST is that increasing numbers of bright Low-Earth-Orbit (LEO) satellites (e.g., Starlink) are being launched, which could leave streaks in astronomical pointings. Over the last three years, many astronomers have raised concerns about the impact of this on the LEO ecosystem and astronomical surveys \citep{lawrence22,tyson20}. LEO satellites are visible from Earth because they reflect sunlight, especially during twilight. As the illuminated LEO satellites move across the field of view of an astronomical pointing, they leave a streak in the image, which negatively impacts the scientific value of the pointing.

XXX--\citet{tyson20} used a very simple algorithm to try and avoid satellite streaks.  In this paper, rather than try to avoid all satellite streaks, we incorporate satellite avoidance as a component on the scheduler's Markov Decision Process, allowing us to avoid a large fraction of satellite streaks without significantly compromising the performance of the survey.

@@

\section{Methodology}
We begin by creating realistic forecasts of three different commercial satellite constellations: StarlinkV1 (4,408 satellites, altitude 540-570 km), OneWeb (6,372 satellites, altitude 1200 km), and Starlink Gen2 (29,988 satellites, altitude 340-614 km). All three of these constellations are highly likely to launch as designed (XXX-lol, no), or close to it, as both operators presently have functional satellites in orbit and are working to build and launch more. These simulated satellite constellations are illustrated in Figure \ref{fig-simulated-constellations}.

% TODO: add three more panels on the top that show each simulated constellation
% as it appeared in the poster
% TODO: make label fonts on each figure GINORMOUS so they are legible when shrunk,
% and delete any extraneous labels
\begin{figure}[ht!]
\epsscale{0.35}
\plotone{plots/starlinkv1.pdf}
\plotone{plots/starlinkv2.pdf}
\plotone{plots/oneweb.pdf}\\
\plotone{plots/sats_slv1.pdf}
\plotone{plots/sats_slv2.pdf}
\plotone{plots/sats_ow.pdf}\\
\plotone{plots/sats_slv1_late.pdf}
\plotone{plots/sats_slv2_late.pdf}
\plotone{plots/sats_ow_late.pdf}
\epsscale{1}
\caption{Hammer projection of the altitude and azimuth positions of our simulated satellite constellations as seen from Rubin Observatory on October 1, 2023 (sun altitude -18 degrees). Red points show un-illuminated satellites, blue points show those illuminated by the sun, and black points are illuminated and above the Rubin altitude limit. The second row is the same, but for 6 hours later when the sun has reached and altitude of -50 degrees. Because the Starlink constellations are lower altitude, they have no illuminated satellites in the middle of the night, while the OneWeb constellation has only one illuminated satellite above the Rubin altitude limit. \label{fig-simulated-constellations} }
\end{figure}

Using these simulated satellite constellations, we add an additional weighting for satellite dodging to the Rubin Scheduler. Given an input satellite distribution, the scheduler predicts satellite maps across a determined time frame. Based on the predicted location of satellites, the scheduler creates a reward weighting: areas with more illuminated satellites will receive a more negative reward weighting, or in other words, the scheduler is incentivized to avoid them. The scheduler then combines the satellite reward weights with other weights to determine where to point the telescope to take the next set of simulated observations. Figure~\ref{fig-simulated-scheduler} shows an example of scheduler weighting for the three simulated satellite constellations at two different sun altitudes.

XXX--quantify the basis function maps a bit more. The positions of the illuminated satellites are computed in 10s intervals. These maps are then summed over 90 minute blocks to generate an avoidance map.

XXX--I'll add some text describing the baseline scheduler that we use--note that we only run for the first year. As a starting point we use the baseline scheduler simulation in \citet{yoachim2022b}.

@@

As shown in Figure \ref{fig-simulated-scheduler}, the predicted OneWeb constellation map has more negative reward weighting than other constellations. Although the OneWeb constellation has fewer satellites than the Starlink Gen2 constellation, the OneWeb constellation is at a higher altitude, meaning that its satellites will be illuminated for a longer portion of the night, and not just close to twilight. Therefore, OneWeb shows more negative weighting than Starlink Gen2 though it has a smaller satellite population. To investigate whether the scheduler behaves how we expect with the new satellite weighting added we create a testing function that measures the length of satellite streaks in the simulated field of views. To ensure efficiency, only satellites that are above the altitude limit and illuminated by the sun will be considered. Satellites below the altitude limit (indicated in the gray region in Figure \ref{fig-simulated-scheduler}) can’t be observed in pointings and are therefore not included for efficiency. For each satellite, we first determine whether they are in the field of view for any given telescope pointing by calculating their distance from the center of the field of view. If their distance from the field of view is less than the radius of the field of view, the satellite has crossed through the pointing. To quantify the impact of the satellite on the pointing, we then project the satellite location as well as the pointing to a 2D x,y plane. In the 2D plane, the field of pointing becomes roughly a circle and the start and end locations of the satellite become two points on the plane. Using the shapely python library, we calculate the intersection length of the satellite location points and the circle. Therefore, with a given simulated satellite constellation and a schedule of observations, we record the number of satellites that occur in each field of view and measure how long the satellite streaks are. Using a realistic estimated satellite streak width of 50 pixels ($\sim10$ arcseconds), we can convert the streak length results into the number of affected image pixels, which allows us to quantitatively measure the efficiency of the dodging algorithm with pixel loss.

@@

\begin{figure}[ht!]
\plottwo{plots/streaklen_v_weight.pdf}{plots/streakfrac_v_weight.pdf}
\plottwo{plots/nexp_v_weight.pdf}{plots/deltam_v_weight.pdf}
\caption{Results from simulating the first year of Rubin observations in the presence of different satellite constellations and varying how strongly the scheduler attempts to avoid satellites.  Top panels show that, as expected, the number of streaks and mean streak length decrease as more emphasis is put on satellite dodging. Bottom panels show the resulting trade off, as avoiding satellites forces the scheduler to make longer slews and observe less desirable parts of the sky, resulting in fewer total exposures in the first year and a shallower median coadded depth in $g$.
xxx--could stack these and share the x-axis. 
\label{fig-pixel-loss-weight}}
\end{figure}

@@

As shown in Figure \ref{fig-trade-off} , for all three constellations, minor reduction in mean pixel loss per visit ($0.001\%$ decrease) results in a significant decrease in median coadded depth. Currently, pixel loss due to satellite streaks only constitutes a very low percentage of total pixels in a pointing (about $0.008\%$ without any dodging). Given that the percentage is currently so small, a very minor reduction in mean pixel loss per visit results in significant reduction in image depth.

\begin{figure}
    \centering
    \plottwo{plots/temp/opsim_Count_night_night_lt_366_and_visitExposureTime_lt_20_HEAL_SkyMap.pdf}{plots/twilight_curves.pdf}
    \caption{Impact of streaks on twilight NEO observations that can not easily be shifted to dodge satellites. Left panel shows the altitude, azimuth distribution of NEO observations in a recent survey simulation. The right panel shows the number of streaks that would result from different satellite constellations as a function of observation altitude. The legend includes an estimate of the pixel loss percentage for NEO observations assuming a 50 pixel wide mask is used. }
    \label{fig:twi_neo}
\end{figure}

@@

We find larger constellations will only leave streaks on $\sim10$\% of Rubin images. This is a much lower value than previous studies \citep{lawrence22, tyson20}. This is partly due to the latest Starlink plans not including higher altitude orbits that were included in the \citet{tyson} study. % And partly due to not conflating twilight observations with the whole survey.

XXX--While satellite streaks are poised to be a nuisance to Rubin observations, it is important to remember 1) They do not saturate detectors, 2) Do not leave streaks when in earth's shadow, 3)

\section{Future Work}
While the satellite dodging algorithm is mostly unnecessary for the three satellite constellations we considered, many astronomers estimate that there will be sharp increases in satellite population in the next 5–10 years, which notably overlaps the LSST operations period (2024–2035). With the dramatic increase in satellite population, the dodging algorithm might become more relevant. In addition, satellites from other operators may be significantly brighter than present-day Starlink and OneWeb satellites, and may saturate the LSST Camera’s detectors. In this scenario, dodging may be required to salvage some science, even if it means losing out on survey depth. Related to this, one future work direction involves adding in brightness weighting to the dodging algorithm. The idea is to only avoid satellites brighter than a certain brightness threshold. This could potentially reduce the region of avoidance, therefore reducing the loss in coadded depth — it could reduce the tradeoff between pixel loss and loss in survey depth, making this preliminary work directly relevant for a successful LSST. Finally, it may be possible to compute optimal starting locations for observation schedules based on predicted satellite maps in order to optimize satellite avoidance.

>>>>

We look at the impact of satellite mega constellations on the planned observations of Rubin Observatory. We construct and simulate the motions of planned Starlink and OneWeb satellites, and test how effectively the Rubin scheduler can avoid satellite streaks when given a satellite forecast. Even in the case of multiple mega constellations in orbit, we expect Rubin would only lose $\sim$0.02\% of science pixels to streaks. Given reasonable satellite orbit predictions we find we can reduce losses by a factor of 2 with minimal impact to the survey science goals. Even so, the tiny fraction of pixels lost makes mitigation strategies unnecessary. Twilight observations at high airmass with no satellite dodging mitigation could see pixel losses up to 0.06\%. 
\end{abstract}

@@

To accomplish this, Rubin observatory has developed a scheduling algorithm that uses a modified Markov Decision Process which can generate lists of desirable observations in real time \citep{naghib19}. The Rubin scheduler balances the desire to minimize slewtime, optimize signal to noise in each image, and to maintain a uniform survey footprint.

One challenge for the LSST is that increasing numbers of bright Low-Earth-Orbit (LEO) satellites (e.g., Starlink) are being launched, which could leave streaks in astronomical pointings. Over the last three years, many astronomers have raised concerns about the impact of this on the LEO ecosystem and astronomical surveys \citep{lawrence22,tyson20}. LEO satellites are visible from Earth because they reflect sunlight, especially during twilight. As the illuminated LEO satellites move across the field of view of an astronomical pointing, they leave a streak in the image. While the flux from satellite streaks can be removed, the resulting pixels typically have such low signal-to-noise as to be scientifically useless.

\citet{tyson20} used a very simple algorithm to see if Rubin could avoid satellite streaks.  In this paper, rather than try to avoid all satellite streaks, we incorporate satellite avoidance as a component on the scheduler's Markov Decision Process, allowing us to avoid a large fraction of satellite streaks without drastically altering the overall performance of the survey.

@@

\section{Methodology}

We begin by creating realistic forecasts of three different commercial satellite constellations: Starlink Gen1 (4,408 satellites, altitude 540-570 km), OneWeb (6,372 satellites, altitude 1200 km), and Starlink Gen2 (29,988 satellites, altitude 340-614 km). OneWeb has successfully launched several hundred satellites, while the number of Starlink satellites is in the thousands. These simulated satellite constellations are illustrated in Figure~\ref{fig-simulated-constellations}.

% TODO: add three more panels on the top that show each simulated constellation
% as it appeared in the poster
% TODO: make label fonts on each figure GINORMOUS so they are legible when shrunk,
% and delete any extraneous labels
\begin{figure}[ht!]
\epsscale{0.35}
\plotone{plots/starlinkv1.pdf}
\plotone{plots/starlinkv2.pdf}
\plotone{plots/oneweb.pdf}\\
\plotone{plots/sats_slv1.pdf}
\plotone{plots/sats_slv2.pdf}
\plotone{plots/sats_ow.pdf}\\
\plotone{plots/sats_slv1_late.pdf}
\plotone{plots/sats_slv2_late.pdf}
\plotone{plots/sats_ow_late.pdf}
\epsscale{1}
\caption{Top row shows the 3-D distribution of our simulated satellite constellations. Middle row shows Hammer projections of the altitude and azimuth positions of satellites as seen from Rubin Observatory on October 1, 2023 (sun altitude -18 degrees). Red points show un-illuminated satellites, blue points show those illuminated by the sun, and black points are both illuminated and above the Rubin 20 degree altitude limit. The bottom row is the same, but for 6 hours later when the sun has reached and altitude of -50 degrees. Because the Starlink constellations are lower altitude orbits, they have no illuminated satellites in the middle of the night, while the OneWeb constellation has only a single illuminated satellite above the Rubin altitude limit at this particular time. \label{fig-simulated-constellations} }
\end{figure}

To simulate Rubin observations, we start with the baseline observing strategy in \citet{yoachim2022b}. This baseline attempts to take most observations in mixed filter pairs (e.g., $r$\ followed by $i$ 33 minutes later), and completes XXX observations in the first year. XXX--some cites for details of the scheduler and model observatory.

The baseline Rubin observing strategy uses three primary basis functions which reward (1) minimizing slewtime, (2) maximizing the depth of images (e.g., by avoiding the moon and high airmass), and (3) maintaining a uniform survey footprint. To this, we add a 4th basis function which rewards avoiding areas of the sky which will have high concentrations of illuminated satellites. Figure~\ref{fig-simulated-scheduler} shows an example of this new basis function for the three simulated satellite constellations at two different sun altitudes. The positions of the illuminated satellites are computed in 10s intervals and marked on the sky. These maps are then summed over 90 minute blocks to generate the basis function maps. Thus our modified scheduler does not try to dodge individual satellite streaks, but rather has a parameterized method for avoiding regions of the sky where satellite streaks are more likely.

@@

As shown in Figure \ref{fig-simulated-scheduler}, the simulated OneWeb constellation has more negative area than other constellations. Although the OneWeb constellation has fewer satellites than the Starlink Gen2 constellation, the OneWeb constellation is at a higher altitude, meaning that its satellites will be illuminated for a longer portion of the night, and not just close to twilight. Therefore, OneWeb shows more negative regions than Starlink Gen2 though it has a smaller satellite population.

To investigate whether the scheduler behaves how we expect with the new satellite weighting added we create a testing function that measures the length of satellite streaks in the simulated field of views. To ensure efficiency, only satellites that are above the altitude limit and illuminated by the sun will be considered. Satellites below the altitude limit (indicated in the gray region in Figure \ref{fig-simulated-scheduler}) can’t be observed in pointings and are therefore not included for efficiency. For each satellite, we first determine whether they are in the field of view for any given telescope pointing by calculating their distance from the center of the field of view. If their distance from the field of view is less than the radius of the field of view, the satellite has crossed through the pointing. To quantify the impact of the satellite on the pointing, we then project the satellite location as well as the pointing to a 2D x,y plane. In the 2D plane, the field of pointing becomes roughly a circle and the start and end locations of the satellite become two points on the plane. Using the shapely python library, we calculate the intersection length of the satellite location points and the circle. Therefore, with a given simulated satellite constellation and a schedule of observations, we record the number of satellites that occur in each field of view and measure how long the satellite streaks are. Using a realistic estimated satellite streak width of 50 pixels ($\sim10$ arcseconds), we can convert the streak length results into the number of affected image pixels, which allows us to quantitatively measure the efficiency of the dodging algorithm with pixel loss.

We simulate observing for only the first year of the planned Rubin 10 year survey. The sruvey strategy does not significantly change in later years, so satellite streak statistics should be similar for the entire survey. We do not include any satellite launches or deorbits, and for simplicity assume the satellite orbital parameters are constant. This should be a fine approximation as long as actual satellite orbital parameters are available $\sim1$\ day in advance so our avoidance basis functions can be constructed. Obviously, if there is no information available on mega constellation orbits, avoiding satellite streaks will be much more difficult.

@@

\begin{figure}[ht!]
\plottwo{plots/streaklen_v_weight.pdf}{plots/streakfrac_v_weight.pdf}
\plottwo{plots/nexp_v_weight.pdf}{plots/deltam_v_weight.pdf}
\caption{Results from simulating the first year of Rubin observations in the presence of different satellite constellations and varying how strongly the scheduler attempts to avoid satellites.  Top panels show that, as expected, the number of streaks and mean streak length decrease as more emphasis is put on satellite dodging. Bottom panels show the resulting trade off, as avoiding satellites forces the scheduler to make longer slews and observe less desirable parts of the sky, resulting in fewer total exposures in the first year and a shallower median coadded depth in $g$.
\label{fig-pixel-loss-weight}}
\end{figure}

@@

As shown in Figure \ref{fig-trade-off}, for all three constellations, minor reduction in mean pixel loss per visit ($0.001\%$ decrease) results in a significant decrease in median coadded depth (0.02-0.2 mags). Currently, pixel loss due to satellite streaks only constitutes a very low percentage of total pixels in a pointing (about $0.008\%$ without any dodging). Given that the percentage is currently so small, a very minor reduction in mean pixel loss per visit results in significant reduction in image depth.

\begin{figure}
    \centering
    % oof, I can't figure out how to force the contour plots to be rasterized. 
    \epsscale{.8}
    \plottwo{plots/temp/opsim_Count_night_night_lt_366_and_visitExposureTime_gt_20_HEAL_SkyMap.pdf}{plots/regular_curves.pdf}
    \plottwo{plots/temp/opsim_Count_night_night_lt_366_and_visitExposureTime_lt_20_HEAL_SkyMap.pdf}{plots/twilight_curves.pdf}
    \epsscale{1}
    \caption{Impact of streaks on a simulation including twilight NEO observations that can not easily be shifted to dodge satellites. Top panels show standard survey observations while bottom panels show results for observations used to search for NEOs. Left panels show the altitude, azimuth distribution of  observations. The right panels show the number of streaks that would result from different satellite constellations as a function of observation altitude. The legends include an estimate of the pixel loss percentage for all observations assuming a 50 pixel wide streak mask. No attempt to dodge satellites was made in these simulations.}
    \label{fig:twi_neo}
\end{figure}

@@

We find larger constellations will only leave streaks on $\sim10$\% of Rubin images. This is a much lower value than previous studies \citep{lawrence22, tyson20}. This is partly due to the latest Starlink plans not including higher altitude orbits that were included in the \citet{tyson20} study. % And partly due to not conflating twilight observations with the whole survey.

While satellite streaks are poised to be a nuisance to Rubin observations, it is important to remember they are not expected to saturate CCD detectors (thus satellite streaks can be removed with relatively narrow pixel masks) and do not leave streaks when in earth's shadow (which is most of the night). Thus, even if there are tens of thousands of satellites, the percentage of pixels lost is incredibly small. For comparison, the Rubin focal plane has a fill factor of XXX\%, meaning XXX\% of potential observations are lost to chip gaps. We also expect diffraction spikes from bright stars to destroy XXX\% of pixels.

\section{Future Work}

XXX-maybe just fold Future Work in with conclusions. I'm not sure we have a ton more future work to do as far as satellite dodging goes.

While the satellite dodging algorithm is mostly unnecessary for the three satellite constellations we considered, many astronomers estimate that there will be sharp increases in satellite population in the next 5–10 years, which notably overlaps the LSST operations period (2024–2035). With the dramatic increase in satellite population, the dodging algorithm might become more relevant. In addition, satellites from other operators may be significantly brighter than present-day Starlink and OneWeb satellites, and may saturate the LSST Camera’s detectors. In this scenario, dodging may be required to salvage some science, even if it means losing out on survey depth. Related to this, one future work direction involves adding in brightness weighting to the dodging algorithm. The idea is to only avoid satellites brighter than a certain brightness threshold. This could potentially reduce the region of avoidance, therefore reducing the loss in coadded depth — it could reduce the tradeoff between pixel loss and loss in survey depth, making this preliminary work directly relevant for a successful LSST. Finally, it may be possible to compute optimal starting locations for observation schedules based on predicted satellite maps in order to optimize satellite avoidance.
